---
title: "Exploratory Data Analysis: Which genes and their expression levels may be associated with developing familial alzheimer's disease?"
author: "Orfeas Gkourlias"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(lemon)
library(knitr)
codebook <- read_delim(file = "codebook.txt",
                       delim = ";",
                       show_col_types=FALSE)
```
\newpage

\section{Introduction}
This document aims to explore, analyse and explain the data set being used
in answering the following research question: "Given 10 attributes, how do they 
compare in predicting the chances of a person's risk of a stroke?". As the res
earch question implies, the data set consists of 10 attributes. This project
has the goal of comparing those attributes, so that the most likely predictors 
for a stroke may be deduced. Some attributes affect each other, while others may
not. Analysis of these correlations can help in finding the rankings of the attr
ibutes.

To get a feel for what the scope and attributes of the data set consists of, it
will be loaded and the first 10 results will be displayed.

```{r}
main <- read.csv("../data/stroke-data.csv")
head(main)
nrow(main)
```

There are 12 attributes, 10 of which will be used in the analysis: Gender, age
hypertension, heart_disease, ever_married, work_type, residence_type, avg_gluco
se_level, bmi and smoking_status. The last column indicates whether the person has 
already experienced a prior stroke. This can be used to the train the machine learning
model which will be utilized to answer the research question.

There are 5110 entries in this data set. This is also why the row numbers will
not be replaced with the id's, because there is no order in the id numbers. They
exceed the number 5110.

The attributes and their units can be seen in the code book on the next page.
\newpage   

\section{Codebook}
```{r}
knitr::kable(codebook)
```
\newpage
\section{Initial Data and Attributes}
In this section, the attributes will be examined individually. What these attri
butes could mean for the research question will be discussed. Correlations will
be observed in a later section. Any preprocessing or cleanup required will
also be performed in this section.

## ID
This column is neither noteworthy for analysis or data structure. This column
will therefore be dropped, because the dataframe used already has row numbers
and this makes the ID redundant.

```{r}
main <- main[2:12]
```

## Age
The age of the patient. At first sight, it might look redundant for this data to
be stored as a float, since most of the data consists of a rounded age number.
Some of the entries contain very young patients. The younger a patient is, the 
more important the specifity of the age is, since the age difference is
still significant at that point. It is for that reason that any patient under the
age of 2 will contain a float number, with two decimal numbers. A couple of those
instances will be shown in vector format below:

```{r}
head(c(main[main$age < 2, 2]))
```
The likelihood of a person experiencing a stroke increases with age. This will
therefore be an important attribute in the analysis.

## Hypertension
This indicates with a 0 or 1 whether the patient is affected by hypertension.
The first attribute which is relevant to the heart status of a patient. These
types of attributes will always be important, because any heart condition tends
to come with an increased risk of experiencing a stroke. Since this is a boolean,
the patient either has hypertension, which is indicated with a 1, or not, which 
is indicated with a 0. This could be a harder type for the later machine learning
model to work with. Correlations will probably be found between hypertension and
the other attributes. No further cleanup is required here.

## Heart Disease
Similar to the previous attribute, this is also an important element when
trying to predict stroke risk. The previous observation also applies to this 
attribute.

## Ever married
This displays whether the patient has ever been married in their lifetime. This
will most likely not bet detrimental in predicting the stroke risks of patients.
But this is part of the dataset, so it will therefore be compared with the other
attributes, to see where it ranks with it's prediction.

## Work Type
A similar attribute to the prior one. Will most likely not be a good predictor
for stroke risk. But it may rank higher than the marriage attribute. Some sector
s could theoretically expose a person to environments where strokes are more likely.

## Residence Type
Considering that some types of residency might be healthier than others, this
attribute may be slightly important in determining the stroke risk of a person.


## Average Glucose Level
This may be more important than the prior three attributes, especially when these
levels are unusually low or high. The literature concerning glucose levels and
their connection to strokes is still being debated. Some papers conclude that
it is not detrimental when observed in non-diabetic people.

## BMI
The body mass index is an indicator for how a person's weight/height ratio. Age 
and gender also being taken into consideration for the calculation. Both extreme
ends of this attribute could be important to the stroke risk of a person. Higher
BMIs are also associated with developing heart disease. Glucose levels may also
be affected. Several other attributes are most likely going to have a correlation
to this attribute.

## Smoking Status
Whether a patient is smoking will most likely affect some of the other attribute
s in this data set. Whether these significant correlations will need to be tested.
The smoking status is unknown for some of the patients.

## Stroke
The column indicating whether the patient has ever had a stroke. This will be
used to train a machine learning model in the actual journal.

\newpage
# Correlations
Following the examination of the intial values and attributes, they may
now be compared so that trends and correlations can be observed. Starting
with the first attribute, which will most likely be important in determining
stroke rist, age. By plotting the summaries of patients who had a stroke, and those
who did not, maybe a link can be seen between the two.

```{r}
no_stroke <- main[main$stroke == 0,]
stroke <- main[main$stroke == 1,]
comb <- rbind(no_stroke, stroke)
ggplot(main, aes(x=factor(stroke), y=age))+geom_boxplot()+ggtitle("Ages by 
                                                                  stroke status")
```
