---
title: "Exploratory Data Analysis: Which genes and their expression levels may be associated with developing familial alzheimer's disease?"
author: "Orfeas Gkourlias"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(lemon)
library(knitr)
codebook <- read_delim(file = "codebook.txt",
                       delim = ";",
                       show_col_types=FALSE)
```

\section{Introduction}
This document aims to explore, analyse and explain the data set being used
in answering the following research question: "Given 10 attributes, how do they 
compare in predicting the chances of a person's risk of a stroke?". As the res
earch question implies, the data set consists of 10 attributes. This project
has the goal of comparing those attributes, so that the most likely predictors 
for a stroke may be deduced. Some attributes affect each other, while others may
not. Analysis of these correlations can help in finding the rankings of the attr
ibutes.

To get a feel for what the scope and attributes of the data set consists of, it
will be loaded and the first 10 results will be displayed.

```{r}
main <- read.csv("../data/stroke-data.csv")
head(main)
nrow(main)
```

There are 12 attributes, 10 of which will be used in the analysis: Gender, age
hypertension, heart_disease, ever_married, work_type, residence_type, avg_gluco
se_level, bmi and smoking_status. The last column indicates whether the person has 
already experienced a prior stroke. This can be used to the train the machine learning
model which will be utilized to answer the research question.

There are 5110 entries in this data set. This is also why the row numbers will
not be replaced with the id's, because there is no order in the id numbers. They
exceed the number 5110.

The attributes and their units can be seen in the code book on the next page.
\newpage   

\section{Codebook}
```{r}
knitr::kable(codebook)
```
\section{Initial Data and Attributes}
In this section, the attributes will be examined individually. What these attri
butes could mean for the research question will be discussed. Correlations will
be observed in a later section. Any preprocessing or cleanup required will
also be performed in this section.

## ID
This column is neither noteworthy for analysis or data structure. This column
will therefore be dropped, because the dataframe used already has row numbers
and this makes the ID redundant.

```{r}
main <- main[2:12]
```

## Age
The age of the patient. At first sight, it might look redundant for this data to
be stored as a float, since most of the data consists of a rounded age number.
Some of the entries contain very young patients. The younger a patient is, the 
more important the specifity of the age is, since the age difference is
still significant at that point. It is for that reason that any patient under the
age of 2 will contain a float number, with two decimal numbers. A couple of those
instances will be shown in vector format below:

```{r}
head(c(main[main$age < 2, 2]))
```
The likelihood of a person experiencing a stroke increases with age. This will
therefore be an important attribute in the analysis.

## Hypertension
This indicates with a 0 or 1 whether the patient is affected by hypertension.
The first attribute which is relevant to the heart status of a patient. These
types of attributes will always be important, because any heart condition tends
to come with an increased risk of experiencing a stroke. Since this is a boolean,
the patient either has hyptertension, which is indicated with a 1, or not, which 
is indicated with a 0. This could be a harder type for the later machine learning
model to work with. Correlations will probably be found between hyptertension and
the other attributes. No further cleanup is requried here.

## Heart Disease
Similar to the previous attribute, this is also an important element when
trying to predict stroke risk. The previous observation also applies to this 
attribute.

## Ever married
