---
title: "Given 10 attributes, how do they compare in predicting the chances of a person's risk of a stroke?"
author: "Orfeas Gkourlias"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(lemon)
library(knitr)
codebook <- read_delim(file = "codebook.txt",
                       delim = ";",
                       show_col_types=FALSE)
main <- read.csv("../data/main.csv", check.names=FALSE)
no_stroke <- main[main$stroke == 0,]
stroke <- main[main$stroke == 1,]
no_s_table <- log(table(no_stroke$heart_disease))
s_table <- log(table(stroke$heart_disease))

vec <- c(no_s_table[1], no_s_table[2], s_table[1], s_table[2])

ratio1 <- as.vector(table(no_stroke$heart_disease)[2])/nrow(no_stroke)
ratio2 <- as.vector(table(stroke$heart_disease)[2])/nrow(stroke)
```
\newpage

\section{Results}
Following the analysis of the dataset and the attributes within, some important
correlations can be observed. Correlations relevant to the research question
mostly consist of heart and health attributes. The first plot regarding heart d
isease differences in non-stroke and stroke patients will be shown.

```{r, echo=F}
barplot(vec, space = c(0,0,1,0), col = c("brown4", "cadetblue"), 
        names.arg = c("                       No Stroke","",
                      "                       Stroke",""), ylab = "Log(Occurance)",
        main = "Heart disease status by stroke history")
legend("topright", c("No Heart disease", "Heart Disease"),
       fill = c("brown4", "cadetblue"))
```

This plot of normalized data shows that there are proportionally more heart disease
patients in the stroke group, than there are in the no stroke group. This ratio
can be translated to a percentage. The percentage of
patients who have not had a stroke is `r round(ratio1*100, 2)`%. This percen
tage is higher for the group of patients who have experienced a stroke. For that
group the percentage is `r round(ratio2*100, 2)`%. This means that it is 4 times as
likely for someone who has had a stroke, to also have a heart condition.
The literature on this subject seems to have come to a common consensus: There is
a correlation between having a heart disease and a higher risk of a stroke, This 
paper published by the American Heart Association being one of the bigger
published works on the topic: [Heart Disease and Stroke Statistics—2020 Update: A Report From the American Heart Association](https://www.ahajournals.org/doi/full/10.1161/CIR.0000000000000757?rfr_dat=cr_pub++0pubmed&url_ver=Z39.88-2003&rfr_id=ori%3Arid%3Acrossref.org)
(Keep in mind, the risk of developing heart disease AFTER  a stroke is also quite big.)

The same can be seen in the case of hypertension, because that is also
a heart condition.

```{r, echo=F}
no_s_table <- log(table(no_stroke$hypertension))
s_table <- log(table(stroke$hypertension))

vec <- c(no_s_table[1], no_s_table[2], s_table[1], s_table[2])

barplot(vec, space = c(0,0,1,0), col = c("brown4", "cadetblue"), 
        names.arg = c("                       No Stroke","",
                      "                       Stroke",""), ylab = "Log(Occurance)",
        main = "Heart disease status by stroke history")
legend("topright", c("No Hypertension", "Hypertension"),
       fill = c("brown4", "cadetblue"))
```
This plot is very similar to the prior one. That makes sense, because quite some
heart conditions are accompanied by hypertension. Hypertension on its own
is already classified as a heart condition too.

\newpage
Age has been observed to also be an important attribute. The older a person beco
mes, the higher the likelihood of a stroke. Especially when combined with other
attributes, such as hypertension or heart conditions. According to this paper, 
the odds of having a stroke double for every ten years a person lives. 
[Aging and ischemic stroke](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6535078/)

```{r}
ggplot(main, aes(x=factor(stroke), y=age))+geom_boxplot()+ggtitle("Ages by stroke status")+
  xlab("Stroke")
```
This shows that the older a person is, the more likely they are to have experienced a stroke.
This would of course make sense, considering that the older someone gets, the 
higher the chance one might encounter health issues.

Another health related observation that may be relevant is that of the BMI means
for stroke and no stroke patients. This can be visualized through a similar box plot
as the prior one.

```{r}
ggplot(main, aes(x=factor(stroke), y=bmi))+geom_boxplot()+ggtitle("BMI by stroke status")+
  xlab("Stroke")
```
The patients who have had a stroke only have a slightly higher BMI. The no stroke
group also has more extreme values on both ends.

```{r}
sapply(stroke[1,], table)
```
