---
title: "Given 10 attributes, how do they compare in predicting the chances of a person's risk of a stroke?"
author: "Orfeas Gkourlias"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
library(tidyr)
library(lemon)
library(knitr)
codebook <- read_delim(file = "codebook.txt",
                       delim = ";",
                       show_col_types=FALSE)
main <- read.csv("../data/main.csv", check.names=FALSE)
no_stroke <- main[main$stroke == 0,]
stroke <- main[main$stroke == 1,]
no_s_table <- log(table(no_stroke$heart_disease))
s_table <- log(table(stroke$heart_disease))

vec <- c(no_s_table[1], no_s_table[2], s_table[1], s_table[2])

ratio1 <- as.vector(table(no_stroke$heart_disease)[2])/nrow(no_stroke)
ratio2 <- as.vector(table(stroke$heart_disease)[2])/nrow(stroke)
```
\newpage

\section{Results}
Following the analysis of the dataset and the attributes within, some important
correlations can be observed. Correlations relevant to the research question
mostly consist of heart and health attributes. The first plot regarding heart d
isease differences in non-stroke and stroke patients will be shown.

```{r, echo=F}
barplot(vec, space = c(0,0,1,0), col = c("brown4", "cadetblue"), 
        names.arg = c("                       No Stroke","",
                      "                       Stroke",""), ylab = "Log(Occurance)",
        main = "Heart disease status by stroke history")
legend("topright", c("No Heart disease", "Heart Disease"),
       fill = c("brown4", "cadetblue"))
```

This plot of normalized data shows that there are proportionally more heart disease
patients in the stroke group, than there are in the no stroke group. This ratio
can be translated to a percentage. The percentage of
patients who have not had a stroke is `r round(ratio1*100, 2)`%. This percen
tage is higher for the group of patients who have experienced a stroke. For that
group the percentage is `r round(ratio2*100, 2)`%. This means that it is 4 times as
likely for someone who has had a stroke, to also have a heart condition.
The literature on this subject seems to have come to a common consensus: There is
a correlation between having a heart disease and a higher risk of a stroke, This 
paper published by the American Heart Association being one of the bigger
published works on the topic: [Heart Disease and Stroke Statistics—2020 Update: A Report From the American Heart Association](https://www.ahajournals.org/doi/full/10.1161/CIR.0000000000000757?rfr_dat=cr_pub++0pubmed&url_ver=Z39.88-2003&rfr_id=ori%3Arid%3Acrossref.org)
(Keep in mind, the risk of developing heart disease AFTER  a stroke is also quite big.)

The same can be seen in the case of hypertension, because that is also
a heart condition.

```{r, echo=F}
no_s_table <- log(table(no_stroke$hypertension))
s_table <- log(table(stroke$hypertension))

vec <- c(no_s_table[1], no_s_table[2], s_table[1], s_table[2])

barplot(vec, space = c(0,0,1,0), col = c("brown4", "cadetblue"), 
        names.arg = c("                       No Stroke","",
                      "                       Stroke",""), ylab = "Log(Occurance)",
        main = "Heart disease status by stroke history")
legend("topright", c("No Hypertension", "Hypertension"),
       fill = c("brown4", "cadetblue"))
```
This plot is very similar to the prior one. That makes sense, because quite some
heart conditions are accompanied by hypertension. Hypertension on its own
is already classified as a heart condition too.

\newpage
Age has been observed to also be an important attribute. The older a person beco
mes, the higher the likelihood of a stroke. Especially when combined with other
attributes, such as hypertension or heart conditions. According to this paper, 
the odds of having a stroke double for every ten years a person lives. 
[Aging and ischemic stroke](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6535078/)

```{r, echo=F}
ggplot(main, aes(x=factor(stroke), y=age))+geom_boxplot()+ggtitle("Ages by stroke status")+
  xlab("Stroke")
```
This shows that the older a person is, the more likely they are to have experienced a stroke.
This would of course make sense, considering that the older someone gets, the 
higher the chance one might encounter health issues.

Another health related observation that may be relevant is that of the BMI means
for stroke and no stroke patients. This can be visualized through a similar box plot
as the prior one.

```{r, echo=F}
ggplot(main, aes(x=factor(stroke), y=bmi))+geom_boxplot()+ggtitle("BMI by stroke status")+
  xlab("Stroke")
```
The patients who have had a stroke only have a slightly higher BMI. The no stroke
group also has more extreme values on both ends.

## Imbalance
Something important which can be observed in the data is the imbalance between
stroke instances. A barplot will be used to demonstrate this imbalance

```{r, echo=F}
ggplot(data = main) +   geom_bar(mapping = aes(x = stroke))+ ggtitle("Stroke 0 to 1 ratio.")
ratio <- round(nrow(main[main$stroke == 0,]) / nrow(main), 3)
```

As can be seen, there is a great imbalance in the amount of patients who
have had a stroke those who have not. To be exact: `r ratio*100`% of the data,
consists of rows where stroke is equal to 0. The remaining 5% has experienced
a stroke. This will be very important when selecting for different machine learn
ing algorithms, because the weighting of occurrences will most likely have to be
changed in this case.

\newpage

# Discussion
Now that the most important results have been shown, some points of contention
could be considered. The aim of this project is to answer the original research
question using machine learning. This may be done with varying algorithms and 
considering different attributes. Looking at the results, some of the attribute
s may seem irrelevant at first, and some would argue these could be removed before
applying any machine learning algorithms. But in this case, none of the attributes
will be discarded. The program being used, Weka, has multiple functions which allow
for the automation of this process. By allowing Weka to select the most influential
attributes systematically, biases may be avoided.

In addition to the data selection, the imbalance is also important to consider.
One way to tackle this problem is by generating samples in the minority class.
That class being the 1 instances under the stroke attribute. Another option was
to use SMOTE (Synthetic Minority Oversampling Technique).  This method was chosen
because it is effectively a more accurate version of oversampling. SMOTE will 
generate synthetic data points, so that the absence of balance will be remedied.
Whether this should be remedied might also be debatable. Some may argue that the
raw data ratio would be an accurate representation. But considering that the 
difference is this extreme, SMOTE was chosen regardless.

## Weighin False positives/ False Negatives
When looking at the significane of false positives and negatives, the effect of
both must be considered relative to their dataset. In this case, false positive
and negatives both have differing degrees of severity. When a false positive gets
detected, a patient will falsely be classified as being at risk of having a stroke.
Using a machine learning model as the only diagnosis method should never
be the only course of action. It is therefore safe to assume that followup
medical examinations of the patient will be performed, allowing for the false p
ositive to be corrected. In this worst case, when the false positive does not
get corrected, a patient may have to adjust their lifestyle and take medication.
There are no negatives to the first solution, since the majority of lifestyle 
adjustments will overlap with the general consensus on what it means to live
a healthy life. The potential of medication would be problematic, because exposing
someone to unnecessary medication is never good.

Now, to compare that to the consequences of a false negative. In the case of a false
negative, a person who is at risk of getting a stroke, will be classified a someone
who is not actually at risk. This means that the person will have to go through life
with an ever increasing risk, since the risk also gets higher the older a person gets.
The implication being that the person will eventually get an unexpected stroke, which
will most likely lead to life debilitating consequences, or even death.

It is for that reason that the false negatives are weighted significantly stricter than
false positives, because the outcomes for a patient are so detrimental for the latter.