---
title: "Results and Discussion & Conlcusion"
output: pdf_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("annotate")
library("hgu133plus2.db")
library("readr")
raw.df = read.csv("../data/GSE28379.csv")
genes <- select(hgu133plus2.db, c(raw.df[,1]), c("SYMBOL","ENTREZID", "GENENAME"))
na.list <- genes[is.na(genes$SYMBOL),] # Storing the genes which were not detected for whatever reason.
colnames(genes)[1] <- "ID_REF" # Renaming  the column so the following merge works.
main <- merge(raw.df, genes, by=c("ID_REF")) # Merging the two dataframes together into a new one.
main <- main[!(main$ID_REF == "!series_matrix_table_end"),] # Removing an indicator row.
main <- main[order(main$log.2.fold.change),] # Reordering the DF by log2FC
high <- main[pnorm(main$log.2.fold.change, mean = mean(main$log.2.fold.change),
                   sd = sd(main$log.2.fold.change)) >= 0.975,]
low <- main[pnorm(main$log.2.fold.change, mean = mean(main$log.2.fold.change),
                  sd = sd(main$log.2.fold.change)) <= 0.025,]
high <- high[order(high$log.2.fold.change, decreasing  = TRUE),]
low <- low[order(low$log.2.fold.change),]
```

## Results
Following the conclusion of exploratory data analysis, multiple important 
observations can be made about the data set. These will be examined by order
of importance to answering the research question.

Starting with the significant fold change values.

```{r echo=F}
head(high[,c(8,10)], 10)
head(low[,c(8,10)], 10)
```
These are the 10 most significant genes, for both the higher and lower ends. There
is a total of 57182 probe reads in the initial dataset. Using the pnorm function
within R, the genes with Log2FC values on both ends were extracted. The higher
end consisting of the upper 2.5% of the distribution, which totals 1761 of the 
57182 genes. The lower 2.5% consists of 1945 genes. These two numbers are close,
reinforcing the conclusion made in  the EDA that the data is normally distributed.
The higher the value, the more up-regulated the gene is in the mutant samples.
The lower values are therefore indicators of down-regulation in the mutant samples.

\newpage
A comparison between the two groups may be made. The following box plots display
summaries of both ends.

```{r echo=F}
boxplot(high$log.2.fold.change, main = "Log2FC Up-Regulation values")
boxplot(low$log.2.fold.change, main = "Log2FC Down-Regulation values")
```
The boxplot shows the usual summary values. It is noteworthy that there are
extreme values in both instances. 6 Of the up-regulated values being higher
than the others in the plot. The same can be seen in the down-regulation plot,
but with 7 values.

```{r echo=F}
d_high <- density(high$log.2.fold.change)
d_low <- density(low$log.2.fold.change)
plot(d_high, main="Density of Up-Regulation", xlab = "Log2FC")
plot(d_low, main="Density of Down-Regulation", xlab = "Log2FC")
```

Density plots also indicate that the down-regulation is on average of bigger effect
than up-regulation. In addition to plotting, this can also be demonstrated with a
numerical summary

```{r echo=F}
print("test")
summary(high$log.2.fold.change)
print("test")
summary(low$log.2.fold.change)
```
As can be seen, the down-regulation has a more extreme mean than that of the
up-regulation.

## Discussion & Conclusion
The examination of the data and plots shows that the mutation indeed affects
several genes. Both in being up and down regulated, with a slight bias towards
down regulation. In answering the research question, there needs to be
a correlation between genes and the development of FAD. This correlation seems
to be present, considering the mutation levels across the three mutation samples
are similar. This may further be researched with addition of other control groups,
and more FAD patient samples.

Which genes are correlated to FAD can also be deduced from the data. There's hundreds
of genes which are expressed significantly different in the mutant samples.
While all of these are important, there's still extremes with an alpha value of
0.05. These extreme down-regulated genes are, in order of significance:
RPS4Y1, DDX3Y, EIF1AY, ZNF257, EIF1AY, HRK and TXLNGY. For up-regulation these are:
BMP5, CASP1, MMP1, ITGB6, LINC02478, CD69.

Looking at the individual genes and their transcripts may point to proteins
which would make sense for FAD patients to be functioning differently.
Looking at RPS4Y1: a gene encoding for  ribosomal protein S4. This specific gene,
and more related to ribosome components, have been observed to be differentialy expressed in
multiple forms of Alzheimers, not just FAD. https://www.sciencedirect.com/science/article/pii/S0197458006004349?via%3Dihub
There's more DEG's which affect the transcriptional capabilities of patients. These have
also been discovered in prior studies. DDX3Y is related to translatoin.
EIF1AY and ZNF257 are both genes encoding for transcription components, 
just like RPS4Y1. (another paper on transcriptional componontent mutation in AD): https://alzres.biomedcentral.com/articles/10.1186/s13195-020-00654-x

The two other up-regulated genes are neither related to the cental dogma of cells.
HRK encodes for a gene related to activation and inhibition of apoptosis. This has also been observed
in prior papers on AD genetics as a possibly important gene for AD ( Hippocampus ). (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6179157/)


TXLNGY seems to be a gene detrimental to syntaxin binding activity. This gene seems to be
differentially expressed in AD patients with choroid plexus epithelium failure.
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4647590/

In the case of down regulation, the BMP5 gene has the lowest log2FC value. This gene
encodes for a protein which is part of developing cartilige and bones. The literature on
this specific gene is quite limited. Cartilige and bone development intuitively does not
allude to any involvement with development of Alzhimer's disease.

CASP1 is a gene that encodes for an enzyme which mediates the cleavage of the inactive precursor of IL-1β.
The literature on the link between this and familial alzheimers disease is a topic of debate. This paper (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2837022/) seems to acknowledge the differential expression of this gene,
but does not accept the association with Alzheimer's disease.

MMP1, ITGB6, LINC02478, CD69.

MMP1 encodes for a metalloproteinase enyzme, which can cause tissue breakdown and secondary damage in diseases. According to this paper, there might be a link with incresaed MMP1 levels and AD. The keyword being elevated. This DEG dataset indicates that the gene is transcribed less in FAD. This is a surprising result, considering what the paper shows.https://pubmed.ncbi.nlm.nih.gov/10984641/

The literatuer on the link between ITGB6 and FAD is limited, but the differential expression of this gene isa recurring osbervation. The protein for which the gene encodes is an integrin.

LINC02478 is a bit of a different gene to the others. This is a long intergenic non-coding rna gene. This might be cause for further analysis, because this gene lacks sufficient research to make any conclusions.

The last gene, CD69, encodes for a membrane-bound, type 2 c-lectin receptor.
https://pubmed.ncbi.nlm.nih.gov/10879686/ Indicates that this is indeed a relevant
DEG. But just like MMP1, there's a difference in up/down regultaion between the data
in the referenced paper, and the dataset used in this one. While the reference shows
that theres a link between up-regulation and AD, the data in this project shows
down-regulation.

It is not clear why the genes in this project seem to have opposing regulation trends
with the data seen in other papers.
